A temporal network analysis of drug co-prescription during antidepressants and anxiolytics dispensing in the Netherlands from 2018 to 2022

### 2025-09-22 docs/submission/GEPI/manuscript/article.tex
Finalize submission to GEPI

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review]{elsarticle}

@@

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{float}
\usepackage{typearea}
\floatplacement{figure}{htbp}
\floatplacement{table}{htbp}
\AtBeginDocument{%
  \storeareas\normalpapersize
}
\BeforeRestoreareas{\cleardoublepage}
\newcommand*\uselandscape{%
  \cleardoublepage
  \KOMAoptions{paper=landscape}%
  \recalctypearea
  \areaset{1.2\textwidth}{1.2\textheight}%
}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A temporal network analysis of drug co-prescription around antidepressants and anxiolytics uses in the Netherlands from 2018 to 2022},
  pdfauthor={Aly Lamuri; Spyros Balafas; Eelko Hak; Jens H. Bos; Frederike Jörg; Talitha L. Feenstra},
  pdfkeywords={Graph Theory, Prescription Registry, Polypharmacy, Mental
Health, Public Health Monitoring},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{A temporal network analysis of drug co-prescription around
antidepressants and anxiolytics uses in the Netherlands from 2018 to
2022}
\author[1,2]{Aly Lamuri%
\corref{cor1}%
}
 \ead{a.l.m.lamuri.murdani@rug.nl} 
\author[1]{Spyros Balafas%
%
}

\author[1]{Eelko Hak%
%
}

\author[1]{Jens H. Bos%
%
}

\author[1,3]{Frederike Jörg%
%
}

\author[1]{Talitha L. Feenstra%
%
}

\affiliation[1]{organization={University of
Groningen},city={Groningen},country={Netherlands},countrysep={,},postcodesep={}}
\affiliation[2]{organization={University of
Indonesia},city={Depok},country={Indonesia},countrysep={,},postcodesep={}}
\affiliation[3]{organization={University Medical Centre
Groningen},city={Groningen},country={Netherlands},countrysep={,},postcodesep={}}

\cortext[cor1]{Corresponding author}

@@

\textbf{Methods:} We utilized the IADB.nl database, including
prescriptions from 128 Dutch pharmacies (2018--2022). A closed cohort of
patients prescribed antidepressants/anxiolytics was extracted.
Medications were classified using the Anatomical Therapeutic Chemical
(ATC) system into 24 therapeutic classes. Time-varying DPNs were
constructed as undirected graphs using symmetric daily dose-adjusted
co-prescriptions. Eigenvector centrality (\(c_e\)) quantified relative
nodal importance. Weekly-aggregated data included prescriptions
(\(n_c\)), number of patients, claim-to-patient ratio, and eigenvector
centrality. Singular spectrum analysis decomposed trends.

\textbf{Results:} Antidepressants (ATC \texttt{N06A}, \(c_e\): 0.09,
\(n_c\): 28,993) and anxiolytics (\texttt{N05B}, \(c_e\): 0.04, \(n_c\):
14,061) had high eigenvector centrality, indicating frequent
co-prescription. Additional high-centrality groups included medications
for alimentary and metabolism (\texttt{A01-A16}), blood
(\texttt{B01-B06}), cardiovascular (\texttt{C01-C10}), analgesics
(\texttt{N02}), and respiratory (\texttt{R01-R07}).

\textbf{Discussion:} DPNs reveal key polypharmacy patterns.
High-centrality medications highlight potential targets for drug
monitoring. This approach aids in identifying influential medications
and refining prescribing oversight.
\end{abstract}

\begin{keyword}
    Graph Theory \sep Prescription
Registry \sep Polypharmacy \sep Mental Health \sep 
    Public Health Monitoring
\end{keyword}
\end{frontmatter}

@@

The complexity of health-related phenomena has led to the growing application of network analysis in medicine and epidemiology \citep{Luke2007,Miglio2021}. Graph theory provides the mathematical foundation for network analysis, offering an approach to study the relationships among the entities of a complex system. The early implementations of network analysis focused primarily on transmission, social, and organizational networks. More recently, \citet{Cavallo2012} introduced drug prescription networks (DPNs) to model population-level drug prescription data. \citet{Bazzoni2015} formalized the term, emphasizing its utility in understanding polypharmacy.

Polypharmacy is defined as a patient with at least two concurrent medications. With a prevalence between 13-90\%, polypharmacy in psychiatry manifests itself in five categories, namely same-class, multiclass, adjunctive, augmentation, and total polypharmacy \citep{Shrivastava2013}. Same-class polypharmacy is the use
of multiple drugs of the same class. Multiclass polypharmacy is
the use of multiple medications of different classes indicated for the
same cluster of symptoms. Adjunctive polypharmacy is the use of additional
medications to treat side effects of other medications. Augmentation
polypharmacy is the use of full-dose and sub-dose medications from a
different class for the same symptom cluster. Finally, total
polypharmacy is the total count of all medications used by a patient, psychopharmaca and other classes as well \citep{Shrivastava2013}. For
policy makers, differences in the definition of polypharmacy pose a challenge in
designing a monitoring system \citep{Sirois2016}. DPN can capture
the complex relationships of a system that can vary over time, which is
very relevant for drugs prescribed for mental disorders.

A DPN for an individual on a given day can be formalized as a simple undirected graph where medications correspond to nodes, and edges connecting pairs of nodes represent co-prescription relationships between medications. Typically, simple undirected graphs are represented by a square symmetric matrix with zeros on the diagonal called the adjacency matrix \citep{estrada2012structure}. In the adjacency matrix of the individual-level DPN, nonzero off-diagonal elements correspond to two medications being co-prescribed on a given day.

@@

Local network statistics, such as node centrality measures, quantify the
importance of a drug in polypharmacy patterns \citep{Askar2021}. Addressing concerns about standardized polypharmacy indicators raised by \citet{Sirois2016} and \citet{Delara2022}, we highlight eigenvector centrality as a key measure for identifying drugs frequently prescribed alongside other influential drugs. Thus, high eigenvector centrality suggests frequent co-prescription with commonly used medications, reflecting co-prescribing trends. Based on previous studies \citep{Cavallo2012, Bazzoni2015}, we hypothesize that eigenvector centrality serves as a meaningful quantitative indicator of polypharmacy. Therefore, this study aims to assess the structural characteristics of temporal DPNs, focusing on the role of eigenvector centrality in identifying influential medications and prescribing patterns in mental health treatments.

@@

\subsection{Source of data}\label{source-of-data}

The data used in this study are exported from IADB.nl, a dynamic pharmacy database containing prescription data from 1994, which continues to grow expansively. The database covers approximately 120 community pharmacies serving over one million patients and was accessed on the 20\textsuperscript{th}of September, 2023. All patients are recorded in the database, regardless of health care insurance, where prescription rates, age, and gender can be generalizable to represent the Netherlands \citep{Visser2013}. The database offers a longitudinal record, where each individual has a uniquely anonymized tracking identifier. All records contain fields on the date of drug dispensing, its quantity, dose regimen, number of days prescribed, prescribing physicians, and the Anatomical Therapeutic Chemical (ATC) code. Except for the over the counter drugs and medication dispensed during hospitalization, other medication records for each patient are complete, mainly due to high patient-pharmacy commitment in the Netherlands. This analysis includes daily medication claims from a static cohort of adults aged 18-65 years old and prescribed for anxiolytics or antidepressants at least once in the time span from 2018 to 2022.

@@

Mathematically, a simple undirected graph \(\mathcal{G}\) is a pair
\(\mathcal{G} = (\mathcal{V}, \mathcal{E})\), where \(\mathcal{V}=\left(1,\dots,n \right)\) is the finite set of nodes and \(\mathcal{E}=\left(\{i,j\} \ |\ i,j\in \mathcal{V},\ i\neq j \right)\) is the finite set of edges with \(\{i,j\}=\{j,i\}\). The adjacency matrix $A\in \{0,1\}^{n\times n}$ of a simple undirected graph \(\mathcal{G}\) with \(|\mathcal{V}|=n\) nodes, is a square symmetric matrix of dimension \(n\times n\) defined as, \[
A_{ij} = 
\begin{cases}
1 & \text{if } \{i, j\} \in \mathcal{E} \\
0 & \text{otherwise}
\end{cases}
\] with \( A_{ij} = A_{ji} \), for all \( i, j \in  \mathcal{V}\) where $i\neq j$. For each day $t$,$t=1,\dots,T$ and each individual $p$, $p=1,\dots,P$ we the adjacency matrix of a simple undirected graph

In a DPN, the
node set
\(\mathcal{V} = \{\mathcal{v}_1, \mathcal{v}_2, \dots, \mathcal{v}_n\}\)
represents a finite set of \(n\) medications. Additionally,
\(\mathcal{E}\) is a finite set of edges, where each edge is a pair of
nodes that corresponds to a co-prescription. In an undirected graph, the
order of \(i, j\) does not matter. Typically, undirected graphs do not
contain self connections,
\(\{\mathcal{v}_i, \mathcal{v}_i\} \notin \mathcal{E}\)
\citep{estrada2012structure}.

A graph can be represented as an adjacency matrix \(A\). The adjacency
matrix \(A\) has dimensions of \(|\mathcal{V}| \times |\mathcal{V}|\),
where \(|\mathcal{V}|\) is the number of nodes in a graph. Each element
of \(a_{ij}\) is the edge, where in an unweighted graph, \(a_{ij} = 1\)
if a connection exists between \(\mathcal{v}_i\) and \(\mathcal{v}_j\),
and \(a_{ij} = 0\) otherwise. In a weighted graph, \(a_{ij}\) can take
any real value (\(a_{ij} \in \mathbb{R}\)) depending on the application
\citep{estrada2012structure}.

\subsection{Data pre-processing}\label{data-pre-processing}

Daily medication claims are represented as tabular data where each row
corresponds to an individual patient's prescriptions. The data was
transformed into a square matrix summarizing co-prescriptions and single
prescriptions to construct a graph-based representation. The rows and
columns of the matrix represent groups of medications, while each matrix
element indicates the frequency of a specific prescription pattern.
Elements in the off-diagonal positions represent co-prescriptions
between two medications, while diagonal elements represent single
prescriptions. This matrix can be interpreted as a graph, where each
medication is a node, and the edges represent co-prescription
frequencies. The weight of the edge is derived from the number of daily
claims. Originally, all defined daily doses (DDD) were assumed to equal
1, leading to straightforward edge weights based on raw counts. However,
since not all medications have DDD values of 1, this assumption could
lead to inaccurate representations. To address this, weights were
adjusted using the density of a normal distribution. The expected value
(\(\mu\)) was set to 1 to prioritize DDD = 1, and the standard deviation
(\(\sigma\)) was set to \(\frac{1}{3}\) to allow for a gradual reduction
in weight as DDD deviates from 1.

\begin{equation}\phantomsection\label{eq-weighted-DDD}{\omega = \frac{1}{\frac{\omega_B}{3} \sqrt{2 \pi}} e^{-\frac{1}{2}\left(\frac{DDD - \omega_B}{\omega_B/3}\right)^2}}\end{equation}

\colorbox{yellow}{What is $\omega_B$ in the formula? Where is $\mu$ that you mention in the text above?}
\colorbox{yellow}{What is the index $B$ in the formula?}\\
\colorbox{yellow}{Why $\mu$ was set to 1? That is the maximum ddd right?}

@@

\colorbox{yellow}{The terms $n_{i,j}$, $\omega_1$, $\omega_2$, etc are not defined somewhere.}\\
\colorbox{yellow}{Having nonzero diagonal is not a usual network representation.}\\
\colorbox{yellow}{What happens to your centrality measures if you set the diagonal equal to zero?}\\
\colorbox{yellow}{I would expect a simpler network structure.}
\begin{center}
    \colorbox{yellow}{Proposal:}
\end{center}
\colorbox{yellow}{For a given day and individual there is an edge with weight equal to 1}\\
\colorbox{yellow}{if there are nonzero ddds in the corresponding pair of drugs.}\\

@@

The choice of centrality largely depends on the objective of network
analysis. As a general guide, degree centrality is useful to identify
popular medication and monitor drug overuse. Betweenness centrality is
suitable for targeting for drug-interaction study and optimizing therapy
plan. Closeness centrality indicates widely-used key medications and
efficiency in treatment networks. Eigenvector centrality is helpful to
identify influential medications and narrow down high-impact medications
for drug monitoring. This study focused on eigenvector centrality to
evaluate which medications have a significant influence on how
antidepressants and anxiolytics are prescribed.
Equation~\ref{eq-eigen-centrality} outlines the calculation of
eigenvector centrality; \(c_i\) and \(c_j\) are the centralities of node
\(\mathcal{v}_i\) and \(\mathcal{v}_j\), respectively; \(\lambda\) is
the eigenvalue; \(a_{ji}\) is an element on row \(j\) and column \(i\)
from an adjacency matrix \(A\), representing the connection between node
\(\mathcal{v}_j\) and \(\mathcal{v}_i\).

\begin{equation}\phantomsection\label{eq-eigen-centrality}{
\displaystyle c_i = \frac{1}{\lambda} \sum_{j \neq i} a_{ji} \cdot c_j
}\end{equation}

@@

The data was extracted as a daily time series and aggregated into weekly
and monthly intervals for analysis. It includes four metrics: daily
medication claims, daily patients, claim-to-patient ratio, and
eigenvector centrality. For each metric, a univariate descriptive
analysis was performed, reporting the mean, median, standard deviation,
and interquartile range to summarize centrality and dispersion.

@@

\subsubsection{Summarizing polypharmacy}\label{summarizing-polypharmacy}

This study explored same-class, multi-class, and total polypharmacy. The
summary statistics were based on prescription-day data aggregated at the
population level over the year. Prescription-day is the number of
medications dispensed on a given day. For example, if seven medications
are prescribed for one day, it accounts for 7 prescription-days. The
mean and standard deviation represent the average number of
prescription-days per person for any polypharmacy regimen.

@@

Since the sum of all eigenvector centralities \(c_i\) in equation
\ref{eq-eigen-centrality} is equal to one, a network with similarly
connected nodes will have an eigenvector centrality \(c_i\) =
\(\frac{1}{n}\) for each node, where \(n\) is the number of nodes. The
value of \(c_i = \frac{1}{n}\) is the expected eigenvector centrality in
a densely connected network where the occurrence of an edge
\(\{\mathcal{v}_i, \mathcal{v}_j\} \in \mathcal{E}\) has the same
probability for all nodes. Per the central limit theorem, the value
\(\frac{1}{n}\) can be parameterized as a mean of a normal distribution,
where all instances of eigenvector centrality will follow the
distribution, \(c_i \sim \mathcal{N}(\mu, \sigma)\). A one-sample
Student's T-test was performed to assess how much each eigenvector
centrality differs from the expected values. As such, eigenvector
centrality greater than \(\frac{1}{n}\) can be considered high, and the
value below it can be considered low. The p-values was corrected for a
type-I error using Bonferroni method.

@@

IADB recorded 149,071 patients having at least one dispensing of
antidepressants or anxiolytics within five years of data extraction.
Table~\ref{tbl-overview-patient} captures the dynamics of the population
from 2018 to 2022. The ratio of male to female only varied slightly, and
the average age steadily increased over the year. These findings imply
that the population demography stays relatively stable overtime without
any indication of sudden changes.

@@

\subsection{Weekly variation of medication
dispensing}\label{weekly-variation-of-medication-dispensing}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  review]{elsarticle}

@@

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{float}
\usepackage{typearea}
\floatplacement{figure}{htbp}
\floatplacement{table}{htbp}
\AtBeginDocument{%
  \storeareas\normalpapersize
}
\BeforeRestoreareas{\cleardoublepage}
\newcommand*\uselandscape{%
  \cleardoublepage
  \KOMAoptions{paper=landscape}%
  \recalctypearea
  \areaset{1.2\textwidth}{1.2\textheight}%
}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Global Epidemiology}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A temporal network analysis of drug co-prescription during antidepressants and anxiolytics dispensing in the Netherlands from 2018 to 2022},
  pdfkeywords={Graph Theory, Prescription
Registry, Co-prescription, Mental Health, Public Health Monitoring},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{A temporal network analysis of drug co-prescription during
antidepressants and anxiolytics dispensing in the Netherlands from 2018
to 2022}

@@

\textbf{Methods:} We utilized the IADB.nl database, including
prescriptions from 128 Dutch pharmacies (2018--2022). A cohort of
patients prescribed antidepressants/anxiolytics was extracted.
Medications were classified using the Anatomical Therapeutic Chemical
(ATC) system into 24 therapeutic classes. Time-varying DPNs were
constructed as undirected graphs using symmetric daily dose-adjusted
co-prescriptions. Eigenvector centrality (\(c_e\)) quantified relative
nodal importance. Weekly-aggregated data included number of dispensing
(\(n_c\)) and eigenvector centrality, which were decomposed using a
singular-spectrum approach.

\textbf{Results:} Antidepressants (\(c_e\): 0.09, \(n_c\): 28,993) and
anxiolytics (\(c_e\): 0.05, \(n_c\): 14,061) had high eigenvector
centrality, demonstrating frequent co-prescription. Other ATC groups
with high centrality included those for the alimentary tract and
metabolism (\texttt{A01-A16}), blood and blood-forming organs
(\texttt{B01-B06}), cardiovascular system (\texttt{C01-C10}),
respiratory system (\texttt{R01-R07}), and analgesics (\texttt{N02}).

\textbf{Discussion:} DPNs revealed key co-prescription patterns.
High-centrality medications highlight potential targets for drug
monitoring, such as identifying co-prescription trends that may warrant
evaluation for safety, appropriateness, or policy oversight. This
approach aids in identifying influential medications and refining
prescribing oversight.
\end{abstract}

\begin{keyword}
    Graph Theory \sep Prescription
Registry \sep Co-prescription \sep Mental Health \sep 
    Public Health Monitoring
\end{keyword}
\end{frontmatter}

@@

The complexity of health-related phenomena has led to the growing
application of network analysis in medicine and epidemiology
\citep{Luke2007, Miglio2021}. Graph theory provides the mathematical
foundation for network analysis, offering an approach to study the
relationships among the entities of a complex system. The early
implementations of network analysis focused primarily on transmission,
social, and organizational networks. More recently, \citet{Cavallo2012}
introduced drug prescription networks (DPNs) to model population-level
drug prescription data. \citet{Bazzoni2015} formalized the term,
emphasizing its utility in understanding polypharmacy.

Psychiatric polypharmacy is defined as a patient with at least two
concurrent psychiatric medications \citep{Shrivastava2013}. With a
prevalence between 13-90\%, psychiatric polypharmacy manifests in five
categories, namely same-class, multi-class, adjunctive, augmentation,
and total polypharmacy \citep{Shrivastava2013}. Same-class polypharmacy
is the use of multiple medications from the same class. Multi-class
polypharmacy is the use of multiple medications from different classes
indicated for the same symptom cluster. Adjunctive polypharmacy is the
use of additional medications to treat side effects due to other
medications. Augmentation polypharmacy is the use of full-dose and
sub-dose medications from a different class for the same symptom
cluster. Finally, total polypharmacy is the overall number of
medications a patient uses, including those prescribed for multiple
psychiatric symptom clusters as well as for comorbid non-psychiatric
conditions \citep{Shrivastava2013}. For policymakers, differences in
defining polypharmacy pose a challenge to design an effective monitoring
system \citep{Sirois2016}. Such monitoring is essential to identify
prevalent co-prescription pattern that may impact treatment
appropriateness and healthcare policy development.

A DPN is a graphical representation where medications correspond to
nodes, and edges represent co-prescription relationships between them.
Typically, simple undirected graphs are represented by a square
symmetric adjacency matrix \(A\), where each nonzero element \(A_{ij}\)
indicates that medications \(i\) and \(j\) were co-prescribed on a given
day. A population-level DPN can be constructed by aggregating individual
patient networks through matrix addition \citep{Cavallo2012}. Analyzing
the structural characteristics of a DPN using both local
(medication-level) and global (network-level) graph measures provides
insights into co-prescription dynamics.

A DPN for an individual on a given day can be formalized as a simple
undirected graph where medications correspond to nodes, and edges
connecting pairs of nodes represent co-prescription relationships
between medications \citep{Cavallo2012}. Typically, simple undirected
graphs are represented by a square symmetric matrix with zeros on the
diagonal called the adjacency matrix \citep{estrada2012structure}. In
the adjacency matrix of the individual-level DPN, nonzero off-diagonal
elements correspond to two medications being co-prescribed on a given
day.

@@

Local network statistics, such as node centrality measures, quantify the
importance of a medication in polypharmacy patterns \citep{Askar2021}.
Addressing concerns about standardized polypharmacy indicators raised by
\citet{Sirois2016} and \citet{Delara2022}, we highlight eigenvector
centrality as a key measure of a medication's connection to highly
prescribed drugs. A high eigenvector centrality suggests frequent
co-prescription with commonly used medications, reflecting prescribing
trends. Building on prior studies \citep{Cavallo2012, Bazzoni2015}, we
hypothesize that eigenvector centrality serves as a meaningful
quantitative indicator of polypharmacy. Therefore, this study aimed to
assess the structural characteristics of temporal DPNs, focusing on the
role of eigenvector centrality in identifying associated medications in
adult patients prescribed for antidepressants or anxiolytics.

@@

\subsection{Source of data and study
population}\label{source-of-data-and-study-population}

All data in this study originated from the University of Groningen
IADB.nl, a dynamic pharmacy database containing prescription data since
1994 which still grows expansively. The database covers approximately
128 community pharmacies serving over one million patients and was
accessed on the 20\textsuperscript{th}of September, 2023. All patients
are recorded in the database, irrespective of health care insurance,
where the prescription rates, age, and gender are generalizable to
represent the Netherlands \citep{Visser2013}. The database offers a
longitudinal record, where each individual has a uniquely anonymized
tracking identifier. All records contain fields on the date of
medication dispensing, its quantity, dose regimen, number of days
prescribed, prescribing physicians, and the Anatomical Therapeutic
Chemical (ATC) code. Except for the over the counter drugs and
medication dispensed during hospitalization, all medication records for
each patient are complete, mainly due to high patient-pharmacy
commitment in the Netherlands.

This analysis includes daily drug administration from a static cohort of
adults aged 18 to 65 years and prescribed anxiolytics or antidepressants
at least once in the period 2018-2022.

@@

Mathematically, a simple undirected graph \(\mathcal{G}\) is a pair
\(\mathcal{G} = \left(\mathcal{V}, \mathcal{E}\right)\), where
\(\mathcal{V} = \left(1, \dots, n \right)\) is the set of nodes and
\(\mathcal{E} = \left(\{i, j\} \ | \ i, j \in \mathcal{V}, \ i \neq j \right)\)
is the finite set of edges. The adjacency matrix
\(A \in \{0, 1\}^{n \times n}\) of a simple undirected graph
\(\mathcal{G}\) is a square symmetric matrix of dimension \(n \times n\)
defined as:

\begin{equation}\phantomsection\label{eq-graph-theory}{
A_{ij} = 
\begin{cases}
1 & \text{if} \{i, j\} \in \mathcal{E} \\
0 & \text{otherwise}
\end{cases}
}\end{equation}

with elements \(A_{ij} = A_{ji}\), for all \(i, j \in \mathcal{V}\)
where \(i \neq j\) \citep{estrada2012structure}.

In our setting, drug co-prescriptions from a single individual \(p\) on
a specific day \(t\) are encoded in a simple undirected graph
\(\mathcal{G}^p(t)=\left(\mathcal{V}, \mathcal{E}^p\left(t\right)\right)\)
with \(n \times n\) adjacency matrix \({A^p}\left(t\right)\). In these
daily observed individual-level networks, the node set \(\mathcal{V}\)
corresponds to a set of \(n\) medication classes according to the ATC
classification, and the edge set \(\mathcal{E}^p\left(t\right)\)
contains edges that indicate the concurrent use of pairs of medications
for a particular individual \(p\) on a given day \(t\).

\subsection{Data pre-processing to build the data
matrix}\label{data-pre-processing-to-build-the-data-matrix}

Assume an observational study where \(P\) individuals indexed by \(p\),
\(p=1,\dots,P\) are followed for each day \(t\), \(t=1, \dots, T\) for a
total period of \(T\) days. To derive the population-level DPN on each
day \(t\) with adjacency matrix \(A_t\), the individual-level daily
prescription matrices \(A^{p}\left(t\right)\) were aggregated by
computing their element-wise sum, that is,

\begin{equation}\phantomsection\label{eq-DPN}{
\displaystyle \sum_{p=1}^P A^{p}\left(t\right), \ \text{where} \ A\left(t\right) \in \ \mathbb{Z}_+^{n\times n}.
}\end{equation}

The matrix \(A\left(t\right)\) captures the population frequency of
co-prescriptions (off-diagonal) and single prescriptions (diagonal) in
the data at

the specified time \(t\). Each element \(A_{ij}\left(t\right)\) in the
matrix was interpreted as the number of patients being prescribed
medication classes \(i\) and \(j\), with \(i,j\in \mathcal{V}\), on the
same day \(t\).

The population-level DPN is then a weighted undirected graph
\(\mathcal{G}\left(t\right) = \left(\mathcal{V}, \mathcal{E}\left(t\right)\right)\)
observed over \(T\) discrete time points \(t=1,\dots,T\), where: -
\(\mathcal{V}\) is the fixed set of \(n\) nodes, -
\(\mathcal{E}_t \subseteq \{ \{i, j\} : i, j \in \mathcal{V}, i \neq j \}\)
is the edge set at time \(t\).

The population-level DPN at time \(t\) is fully described by its
symmetric weighted adjacency matrix
\(A\left(t\right) \in \mathbb{Z}_+^{n \times n}\), which is defined as:

\[
{A}_{ij}\left(t\right) =
\begin{cases}
w_{t,ij} > 0 & \text{if } \{i, j\} \in \mathcal{E}\left(t\right), \\
0 & \text{otherwise},
\end{cases}
\quad \text{with } {A}\left(t\right) = {{A}\left(t\right)}^\top.
\]

In the original work of DPN, all defined daily dose (DDD) were assumed
equal to 1, allowing for applying raw counts as the edge weights.
However, since not all medications have DDD = 1, this assumption could
lead to inaccurate representations. To enhance the accuracy of edge
weights, we introduced a weighting function to adjust the contribution
of each co-prescription. Specifically, each co-prescribed medication
class was assigned a weight \(\omega_j\) based on its DDD value using a
Gaussian kernel centered at the baseline weight \(\omega_B = 1\), i.e.,
the expected DDD, with standard deviation \(\sigma = \frac{1}{3}\) to
allow for a gradual reduction in weight as DDD deviates from 1.

\begin{equation}\phantomsection\label{eq-weighted-DDD}{\omega_j = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{\left(DDD_j - 1\right)}{2\sigma^2}^2}}\end{equation}

The edge weight between medication classes \(i\) and \(j\) was computed
as the average of their individual weights:

@@

This weighting approach scaled the edge weight that deviate
substantially from the standard \(\omega_B = 1\), improving
comparability across medication classes. To remain consistent with
standard network representations and avoid confounding self-loops, the
diagonal elements were omitted.

@@

The choice of centrality largely depends on the objective of network
analysis. As a general guide, degree centrality is useful to identify
popular medication and monitor drug overuse. Betweenness centrality is
suitable for targeting for drug-interaction study and optimizing therapy
plan. Closeness centrality indicates widely-used key medications and
efficiency in treatment networks. Eigenvector centrality is helpful to
identify influential medications and narrow down high-impact medications
for drug monitoring. We defined high and low eigenvector centrality
relative to the expected value \(\frac{1}{n}\), based on the null model
of uniform connectivity (see ``Determining relative importance''). This
study focused on eigenvector centrality to evaluate which medications
have a significant influence on the prescription of antidepressants and
anxiolytics. Equation~\ref{eq-eigen-centrality} outlines the calculation
of eigenvector centrality; \(c_i\) and \(c_j\) are the centralities of
nodes \(\mathcal{v}_i\) and \(\mathcal{v}_j\), respectively, with
\(i, j \in \mathcal{V}\); \(\lambda\) is the eigenvalue;
\(A_{ji}\left(t\right)\) is an element on row \(j\) and column \(i\)
from an adjacency matrix \(A\left(t\right)\), representing the
connection between node \(\mathcal{v}_j\) and \(\mathcal{v}_i\) with
\(i, j \in \mathcal{V}\).

\begin{equation}\phantomsection\label{eq-eigen-centrality}{
\displaystyle c_i\left(t\right) = \frac{1}{\lambda} \sum_{j \neq i} A_{ji}\left(t\right) \cdot c_j\left(t\right)
}\end{equation}

@@

The data was extracted as a daily time series and aggregated into weekly
and monthly intervals for analysis. It includes four metrics: daily
prescription dispensing, daily patients, dispensing-to-patient ratio,
and eigenvector centrality. For each metric, a univariate descriptive
analysis was performed, reporting the mean, median, standard deviation,
and interquartile range to summarize centrality and dispersion.

@@

\subsubsection{Summarizing
co-prescription}\label{summarizing-co-prescription}

This study explored same-class, multi-class, and total co-prescription
categories. These three groups were selected as they can be directly
inferred from medication dispensing data aggregated at the population
level over the year. The statistics were summarized as prescription-day,
defined as the number of medications dispensed on a given day. For
example, seven medications prescribed for one day account for 7
prescription-days. The mean and standard deviation represented the
average number of prescription-days per person for any co-prescription
regimen.

Although five polypharmacy categories were described by
\citet{Shrivastava2013}, augmentation and adjunctive polypharmacy
require clinical context such as diagnoses to determine the rationale
for combined prescriptions, which is not available in our data source.
Therefore, focusing on the three categories that rely solely on
medication class information is the most reasonable approach for this
study.

@@

Since the sum of all eigenvector centralities \(c_i\) in equation
\ref{eq-eigen-centrality} equals 1, the expected eigenvector centrality
for each node in a uniformly connected network is \(\frac{1}{n}\), where
\(n\) is the total number of nodes. In such a network, each node has an
equal probability of being connected to any other, and thus no node is
more ``influential'' than another. This expected value value,
\(\frac{1}{n}\), serves as a baseline under a null model of uniform
connectivity. Assuming eigenvector centrality scores approximately
follow a normal distribution, we applied a one-sample Student's T-test
with Bonferroni correction to assess how much each eigenvector
centrality differs from the expected values. Nodes with \(c_i\) greater
than \(\frac{1}{n}\) were categorized as having high eigenvector
centrality, and those below the expected value were considered low.

@@

IADB recorded 149,071 patients having at least one dispensing of
antidepressants or anxiolytics within five years of data extraction.
Table~\ref{tbl-overview-patient} captures the demographical dynamics of
the population from 2018 to 2022. The ratio of male to female only
varied slightly, and the average age steadily increased over the year.
These findings imply that the population demography stays relatively
stable overtime without any indication of sudden changes.

@@

\subsection{Cyclicality and seasonality of medication
dispensing}\label{cyclicality-and-seasonality-of-medication-dispensing}
